### 2018-04-23 contents/reviews/pain.tex

The earliest theory that attributes pain to a dedicated physical pathway is commonly credited to Ren\'{e} Descartes (\ref{fig:descarte}). However, for Descartes, pain as an experience  could not be divorced from the idea of the soul, as he had stated in his letter to the French polymath Marin Mersenne: "I do not explain the feeling of pain without reference to the soul. For in my view exists only in the understanding. What I do explain is all the external movements which accompany this feeling in us." \cite[p. 148]{BookCottingham1991v3}. In his \textit{Treatise of Man}, published in 1664, he stated: "Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch. In this way they pull the tiny fibre which you see attach to it, and simultaneously open the entrance to the pore, located opposite the point where this fibre terminates -- just as when you pull one of a string, you cause a bell hanging at the other end to ring at the same time. When the entrance to the pore or small tube is opened in this way, the animal spirits from cavity enter and are carried through it...the fibres cause a movement in the brain which gives occasion for the soul...to have the sensation of pain" \cite[p.101-103]{BookCottingham1991v1}.

@@

The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

 \begin{figure}[ht]
 \includegraphics[width=\textwidth]{gate-control-diagram.pdf}
 \centering
 \caption{Melzack's gate control theory diagram.}
 \label{fig:gate-theory} 
 \end{figure}

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes the necessary information for pain. That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities constitute the sensory component of the pain experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn (laminae V), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzack, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

@@

\textit{Nociceptors}, the receptors in the peripheral nervous system that detect tissue damage can respond to a range of electrical or chemical signals released by noxious stimuli on the tissue, such as physical trauma or heat/cold temperatures exceeding the pain threshold. These receptors broadly separate into two categories, differentiated by the type of axonal innervation they receive. Acute pain receptors are innervated by $A\delta$ fibres, which permits fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma. Dull pain is triggered by receptors that are innervated by slow conducting C fibres ($<1$ m/s). $A\delta$ and C fibres can be co-activated in the presence of noxious stimuli.

@@

The peripheral nociception signals are conducted from the axonal endings of the nociciption neurons, whose cell bodies are located in the spinal cord dorsal root gangions, to the dorsal horn of the spinal cord (\ref{fig:drg}). The $A\delta$ and C afferents project directly to lamina I of the dorsal horn. $A\delta$ afferents also directly project to lamina V. Lamina II, or the substantia gelatinosa, also receives inputs from C afferents, which indirectly synapse to lamina I neurons through interneurons. The dorsal horn also contains a dense layer of local interneurons in lamina III and IV, and receive input from innocuous somatosensory afferents at these layers.

@@

Another spinothalamic termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse. The MD nucleus projects third order afferents to the anterior cingulate cortex (ACC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.

@@

Pain percept, defined as the totality of the experience of pain, is thought to involve distributed regions of the brain, coined as the “pain neuromatrix”\cite{Melzack1999a}. The regions are not functionally independent, but rather thought to be dynamically inter-related.

 Important cortical regions involved in pain include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). The role of these cortical regions in pain were studied in neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during noxious stimulation. (Figure \ref{fig:pain-matrix}).

@@

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain processing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions. The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the anterior insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}. The unpleasantness of pain has been thought to be a motivational force \cite{Auvray2010}, that biases actions towards avoidance. There is evidence that the motor cortex, which is involved in action planning, and the ACC are activated during pain \cite{Erpelding2013}. The cortical excitability of the motor cortex is also shown to be altered in chronic pain \cite{Parker2016}. The motivational system is also involved in pain-reward and involves the ACC, lateral orbitofrontal cortex (OFC), and the prefrontal cortex (PFC, both ventromedial and anterior aspects) \cite{Navratilova2015}. The nucleus accumbens (NAc) BOLD activity correlated with anticipation of pain relief \cite{Baliki2010}. The reward/motivational circuitry is thought to be disrupted in chronic pain, and such disruption is related to the emotional and cognitive deficits in chronic pain sufferers.

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation. Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. DMN activation has been correlated with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission, processing, and modulatory circuitry. 
Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex. The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation.

@@

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis, such that the synchronicity of action potential conduction between cortical and subcortical structures are disrupted, and thereby give rise to neuropathic pain. It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain \cite{Geha2008}.

>>>>

The earliest theory that attributes pain to a dedicated physical pathway is commonly credited to Ren\'{e} Descartes (\ref{fig:descarte}). However, for Descartes, pain as an experience  could not be divorced from the idea of the soul, as he had stated in his letter to the French polymath Marin Mersenne: "I do not explain the feeling of pain without reference to the soul. For in my view exists only in the understanding. What I do explain is all the external movements which accompany this feeling in us." \cite[p. 148]{BookCottingham1991v3}. In his \textit{Treatise of Man}, published in 1664, he stated: 
\begin{quote}"Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch. In this way they pull the tiny fibre which you see attach to it, and simultaneously open the entrance to the pore, located opposite the point where this fibre terminates -- just as when you pull one of a string, you cause a bell hanging at the other end to ring at the same time. When the entrance to the pore or small tube is opened in this way, the animal spirits from cavity enter and are carried through it...the fibres cause a movement in the brain which gives occasion for the soul...to have the sensation of pain" \cite[p.101-103]{BookCottingham1991v1}.
\end{quote}

@@

The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells in the spinal dorsal horn that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

 \begin{figure}[ht]
 \includegraphics[width=\textwidth]{gate-control-diagram.pdf}
 \centering
 \caption{Melzack's gate control theory diagram. Modified from \protect\cite{Melzack1965a}}
 \label{fig:gate-theory} 
 \end{figure}

An important contribution of the gate-control theory was to emphasize central nervous system mechanisms \cite{Melzack1999}, and the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. Contemporary views consider pain to be complex and multi-dimensional, where the processing of ascending nociceptive signals, in conjunction with descending modulatory pathways, shape the pain experience. Melzack and Casey further proposed that pain consists of sensory-discriminative, affective-motivational, and cognitive-evaluative components \cite{Melzack1968}. Specifically, the sensory-discriminative component consists of peripheral nociceptive afferents transmit to the spinal dorsal horn (laminae V), whereby nociceptive specific (NS) and wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzack, is the  neuromatrix theory of pain, which dictates that the pain experience is produced by a "neurosignature" pattern, that is generated from widely distributed neural network, which Melzack refers as the "body-self neuromatrix". \cite{Melzack1999}. The concept is now more commonly referred to as the "pain matrix" \cite{tracey2007a}. (See section \ref{section:neuromatrix})

@@

\textit{Nociceptors}, the peripheral nociceptors respond to noxious stimuli. These receptors have free nerve endings, and can respond to thermal, mechanical, and/or chemical stimuli, such as physical trauma or heat/cold temperatures. These receptors broadly separate into two categories, differentiated by the fibre type. Acute nociceptors are associated with $A\delta$ fibres, which permit fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma. Dull pain is triggered by receptors that are associated with slow conducting C fibres ($<1$ m/s). $A\delta$ and C fibres can be co-activated in the presence of noxious stimuli. C-fibre nociceptors can be further differentiated into mechano-heat, silent, and polymodal subtypes. $A\delta$-fibre nociceptors can be divided into high-threshold mechano-heat, low-threshold mechano-heat, and noxious cold subtypes.

@@

The peripheral nociception signals are conducted from the axonal endings of the nociception neurons, whose cell bodies are located in the spinal cord dorsal root gangions, to the dorsal horn of the spinal cord (\ref{fig:drg}). The $A\delta$ and C afferents project directly to lamina I of the dorsal horn. $A\delta$ afferents also directly project to lamina V. Lamina II, or the substantia gelatinosa, also receives inputs from C afferents, which indirectly synapse to lamina I neurons through interneurons. The dorsal horn also contains a dense layer of local interneurons in lamina III and IV, and receive input from innocuous somatosensory afferents at these layers.

@@

Another spinothalamic termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse. The MD nucleus projects third order afferents to the anterior/middle cingulate cortex (ACC/MCC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.

@@

Pain percept, defined as the subjective experience of pain, is thought to involve distributed regions of the brain, coined as the pain “neuromatrix”\cite{Melzack1999a}, or more commonly the "pain matrix". The regions are not functionally independent, but rather thought to be dynamically inter-related.

 Important cortical regions involved in pain include anterior and middle cingulate cortices (ACC/MCC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). The role of these cortical regions in pain were studied in neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during noxious stimulation. (Figure \ref{fig:pain-matrix}).

@@

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain processing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions. The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the anterior insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}. The unpleasantness of pain has been thought to be a motivational force \cite{Auvray2010}, that biases actions towards avoidance. There is evidence that the motor cortex, which is involved in action planning, and the ACC/MCC are activated during pain \cite{Parker2016,Burns2016,Erpelding2013}. The cortical excitability of the motor cortex is also shown to be altered in chronic pain \cite{Parker2016}. The motivational system is also involved in pain-reward and involves the ACC, lateral orbitofrontal cortex (OFC), and the prefrontal cortex (PFC, both ventromedial and anterior aspects) \cite{Navratilova2015}. The nucleus accumbens (NAc) BOLD activity correlated with anticipation of pain relief \cite{Baliki2010}. The reward/motivational circuitry is thought to be disrupted in chronic pain, and such disruption is related to the emotional and cognitive deficits in chronic pain sufferers.

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation. Regions such as vmPFC, dMPFC and the anterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. Pain related DMN deactivation have been found to be attenuated  in mind wandering away from pain, and may reflect antinociceptive processes \cite{Kucyi2013}, and that the DMN may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission, processing, and descending modulatory circuitry. 
Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex. The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation.

@@

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis. It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain \cite{Geha2008}.

### 2018-04-23 contents/reviews/trigeminal.tex

The afferent branches converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion), which then forms the trunk or cisternal component of CN V. The TRG contains all the trigeminal sensory cell bodies. It situates in the Meckel's cave, in the middle cranial fossa. It is anatomically organized with respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.

The extension of the posterior aspect of the TRG forms the trigeminal nerve body, with both sensory and motor roots. The nerve body straddles the TRG and the trigeminal nucleus, by crossing the pontine cistern. The portion of the sensory nerve that enters the pons is termed the trigeminal root entry zone (REZ). The cistern trigeminal sensory nerve, similar to other peripheral nerves, is myelinated by Schwann cells. The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons. The CNS myelin can be several millimetres beyond the surface of the pons, before giving away to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}. A study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm). The transition zone can be up to 25\% of the nerve length \cite{Peker2006}.

@@

CN V passes into the pons (\ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem. It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem. The trigeminal brainstem sensory nucleus complex is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei. The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.

>>>>

The afferent branches converge and form the trigeminal root ganglion (TRG) (also known as Gasserian, or semilunar ganglion), which then leads to the trunk or cisternal component of CN V. The TRG contains all the trigeminal sensory cell bodies. It situates in the Meckel's cave, in the middle cranial fossa. It is anatomically organized with respect to the trigeminal branches, with V1 situated in medial anterior, V2 in caudal lateral, and V3 in the middle.

The extension of the posterior aspect of the TRG forms the trigeminal nerve body, with both sensory and motor roots. The nerve body straddles the TRG and the trigeminal nucleus, by crossing the pontine cistern. The portion of the sensory nerve that enters the pons is termed the trigeminal root entry zone (REZ). The cistern trigeminal sensory nerve, similar to other peripheral nerves, is myelinated by Schwann cells. The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons. The CNS myelin can be several millimetres beyond the surface of the pons, before leading to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}. A study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm). The transition zone can be up to 25\% of the nerve length \cite{Peker2006}.

@@

CN V passes into the pons (Figure \ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem. It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem. The trigeminal brainstem sensory nucleus complex is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei. The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.
